Program Transformations for miniKanren

### 2020-07-05 papers/aplas/text/evaluation.tex
Evaluation: double appendo

\section{Evaluation}
In our study we compared the new non-conjunctive partial deduction with the \ecce partial deduction system.
\ecce is designed for \pro programming language and cannot be directly applied to \mk.
To be able to compare our approach with \ecce, we converted each input program first to the pure subset of \pro, then transformed it with \ecce, and then we converted the result back to the pure subset of \pro.
The coversion to \pro is a simple syntactic conversion.
In conversion from \pro to \mk, a conjunction is generated by each horn clause in which unifications are placed before any relation call.
This mimics in \mk the execution order of \pro program within a conjunction.

@@

The tables and graphs use the following denotations.
\emph{Original} represents the execution time of a program before any transformations were applied; \emph{ECCE} --- of the program transformed with \ecce; \emph{Non-CPD} --- of the program transformed by our approach.
The first two examples also contain \emph{Ideal} row which contain the execution time of the ideal implementations of the relations presented in the paper~\cite{de1999conjunctive}.

@@

The relation \lstinline{doubleAppend$^o$} concatenates three lists by a conjunction of two calls of the \lstinline{append$^o$} relation (see fig.~\ref{doubleApp}).
These two calls share a variable --- an intermediate list \lstinline{ts} --- which should be treated with care during specialization.
Partial deduction does ignore the sharing of the variables which renders the approach to be unproductive for this kind of relations.
This is why this relation is particularly well known in the conjunctive partial deduction work.

The first list gets traversed twice to construct the result.
First, when the intermidiate list \lstinline{ts} is constructed during the first \lstinline{append$^o$} call and then, when \lstinline{ts} is read during the second call.
This double traversal negatively impacts the execution time.

@@

The better implementation of this relation (see fig.~\ref{doubleApp:ideal}) does not traverse the first list twice.
This ideal implementation is generated with the conjunctive partial deduction.

@@

We run all translated programs in the forward and the backward direction.
The forward direction \lstinline{doubleAppend$^o$ x y z ?} concatenates three given lists: the first one of length 700, the second and the third --- of length 1.
When run in the backward direction \lstinline{doubleAppend$^o$ ? ? ? r} , it searches for the first 100 list triples whose concatenation gives the given list of length 700.

There is no significant difference in the execution time in the backward direction between different specialized versions~\ref{tbl:doubleApp}.
However the execution time in the backward direction differs: the program generated by \ecce shows the same speedup as the ideal program, while non-conjunctive partial deduction slows the program down.
\todo{WHY?}

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
                   & forward & backward \\
  \hline\hline
  Original         & 0.0040s & 0.0195s \\ \hline
  \ecce             & 0.0031s & 0.0201s \\ \hline
  Non-CPD          & 0.0055s & 0.0202s \\ \hline
  Ideal           & 0.0031s & 0.0203s \\
  \end{tabular}
  \caption{Evaluation results for doubleAppendo}
  \label{tbl:doubleApp}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{c||c}
                   & [1..100] \\ \hline\hline
  Original         & 2.286ms  \\ \hline
  Ideal           & 4.623ms  \\ \hline
  Ideal removed   & 3.133ms  \\ \hline
  \ecce             & 1.701ms  \\ \hline
  Non-CPD          & 2.578ms
  \end{tabular}

@@

Surprising enough, the execution time of the ideal program is worse than of the original.
\ecce succeeds at improving the program performance, while non-conjunctive partial deduction worsens it a little.
\todo{why}

@@

The non-conjunctive partial deduction first unfolds those calls which are selected with the heuristic.
Since exploring boolean operations makes more sense, they are unfolded before recursive calls of \lstinline{eval$^o$}.
The way non-conjunctive partial deduction treats this program is the same as it treats the other implementation in which boolean operations are moved to the left, as shown in listing~\ref{eval:fst}.
This program is easier for CPD to transform which demonstrates how unequal is the behaviour of CPD for similar programs.

@@

The other way to implement boolean operations is via one basic boolean relation such as \lstinline{nand$^o$} which is, in turn, has a table-based implementation (see listing~\ref{not:nando}).
It will take several sequential unfoldings to derive that variables \lstinline{v} and \lstinline{w} should be \lstinline{^true} when considering a call \lstinline{and$^o$ v w ^true} implemented via a basic relation.
Non-conjunctive partial deduction drives the selected call until it derives useful substitutions for the variables involved while CPD with deterministic unfolding may fail to do so.

@@

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
                   & last    & plain  \\
  \hline\hline
  Original         & 0.695s  & 1.426s \\
  \hline
  \ecce             & 0.205s  & 0.264s \\
  \hline
  Non-CPD          & 0.132s  & 0.137s \\
  \end{tabular}

@@

The \lstinline{unify} relation was executed to find a unifier of two terms $f(X, X, g(Z, t))$ and $f(g(p, L), Y, Y)$.
The original \mk program fails to terminate on this goal in 30 seconds.
On this example, the most performant is the program generated by CPD (0.352 seconds) while the program generated by adding branching heuristic also fails to terminate in 30 seconds.
The non-conjunctive partial deduction shows some improvement with the residual program terminated within 1.947 seconds.
While driving this program, the non-conjunctive partial deduction does too much unfolding which negatively impacts the running time as compared to CPD.

The last test executed \lstinline{isPath} relation to search for 3 paths of length 7 in the graph with 20 vertices and 30 edges.
On this program, non-conjunctive partial deduction showed better transformation results than CPD, although the difference is not that drastic.

All evaluation results are presented in the table~\ref{tbl:unify}.
Each column corresponds to the relation being run as described above.
The row marked ``Original'' contains the running time of the original \mk relation before specialization, ``CPD'' and ``Non-CPD'' correspond to conjunctive and non-conjunctive partial deduction respectively while ``Branching'' is for the CPD modified with the branching heuristic.

\begin{table}
  \centering
  \begin{tabular}{c||c|c|c}
    \multirow{ 2}{*}{Terms} &
    f(X, a) & f(a \% b \% nil, c \% d \% nil, L) & f(X, X, g(Z, t))  \\
    \cline{2-4} &
    f(a, X) & f(X \% XS, YS, X \% ZS) & f(g(p, L), Y, Y)  \\
    \hline\hline
  Original         & $0.247*10^{-4}s$ & $>30s$  & $>30s$   \\
  \hline
  Non-CPD          & $0.084*10^{-4}s$ & $0.021s$ & $2.198s$  \\
  \hline
  \ecce             & $0.128*10^{-4}s$ & $0.026s$ & $1.340s$ \\
  \hline
  \end{tabular}

@@

% \begin{table}
%   \centering
%   \begin{tabular}{c||c|c|c||c}
%                    & \multicolumn{3}{c||}{unify} & isPath \\
%   \hline
%                    & first  & second & third     &        \\
%   \hline\hline
%   Original         & $0.247*10^{-4}s$ & $>30s$  & $>30s$  & $23.743s$ \\
%   \hline
%   Non-CPD          & $0.084*10^{-4}s$ & $0.021s$ & $2.198s$  & $1.786s$  \\
%   \hline
%   \ecce             & $0.128*10^{-4}s$ & $0.026s$ & $1.340s$  & $1.292s$ \\
%   \hline
%   \end{tabular}

@@

We compared two possible implementations of these typing rules.
The first one is obtained by unnesting of the functional program as described in~\cite{lozov2019relational}.
The second program is hand-written in \oc.
Each implementation has been transformed with non-conjunctive partial deducer and by \ecce.

We measured the time needed to generate 100 terms of type integer (averaged over 10 iterations).
\ecce significantly worsens the execution time for both implementations.
Non-conjunctive partial deduction improves the first implementation a little bit, while worsening the performance of the second implementation 3 times.

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
              & Implementation 1 & Implementation 2 \\ \hline\hline
  Original    & 0.464s           & 0.057s           \\ \hline
  Non-CPD     & 0.448s           & 0.154s           \\ \hline
  \ecce        & 1.518s           & 2.543s           \\
  \end{tabular}

>>>>

\section{Evaluation}
In our study we compared the new conservative partial deduction with the \ecce partial deduction system.
\ecce is designed for \pro programming language and cannot be directly applied to \mk.
To be able to compare our approach with \ecce, we converted each input program first to the pure subset of \pro, then transformed it with \ecce, and then we converted the result back to \mk.
The coversion to \pro is a simple syntactic conversion.
In conversion from \pro to \mk, a conjunction is generated by each horn clause in which unifications are placed before any relation call.
This mimics in \mk the execution order of \pro program within a conjunction.

@@

The tables and graphs use the following denotations.
\emph{Original} represents the execution time of a program before any transformations were applied; \emph{ECCE} --- of the program transformed with \ecce with default conjunctive control setting; \emph{ConsCPD} --- of the program transformed by our approach.
The first two examples also contain \emph{Ideal} row which contain the execution time of the ideal implementations of the relations presented in the paper~\cite{de1999conjunctive}.

@@

The relation \lstinline{doubleAppend$^o$} concatenates three lists by a conjunction of two calls of the \lstinline{append$^o$} relation (see fig.~\ref{doubleApp}).
These two calls share a variable --- an intermediate list \lstinline{ts}.
The list \lstinline{xs} is traversed twice to construct the result: first when \lstinline{ts} is constructed and then, when \lstinline{ts} is traversed during the second call.
This double traversal negatively impacts the execution time, but it can be removed by deforestation.

@@

The better implementation of this relation (see fig.~\ref{doubleApp:ideal}) does not traverse the first list twice.
It first recursively copies \lstinline{xs} into the beginning of \lstinline{res}, and only when \lstinline{xs} is fully consumed, it calls the \lstinline{append$^o$} to concatenate \lstinline{ys} and \lstinline{zs}.
This implementation is provided as the ideal implementation of \lstinline{doubleAppend$^o$} in~\cite{de1999conjunctive}.

@@

To automatically transform the initial program into the ideal implementation, the transformer should treat the conjuncts which share variables with care.
Partial deduction splits any conjunction into its subconjunctions, regardless variable sharing, while CPD only splits conjunctions when potential non-termination is detected.
Conservative partial deduction processes subconjunctions in isolation and then joins them back together, if it can help to restrict the search space.
In this example, ConsPD fails to perform deforestation: ConsPD first splits the conjunction \lstinline{append$^o$ xs ys ts /\ append$^o$ ts zs rs} into two relation calls of \lstinline{append$^o$}, then unfolds the first call, but since the second call is a renaming of the first one, it is not unfolded.
The generated program is almost the same as the original: the only difference being that the fresh variables are introduced on the toplevel.

We run all programs in the forward and the backward direction.
The forward direction \lstinline{doubleAppend$^o$ x y z ?} concatenates three given lists: the first one of length 700, the second and the third --- of length 1.
The backward direction \lstinline{doubleAppend$^o$ ? ? ? r} searches for the first 100 list triples whose concatenation gives the given list of length 700.

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
                  & forward & backward \\
  \hline\hline
  Original        & 34.8ms & 1.6ms \\ \hline
  \ecce           & 26.7ms & 1.7ms \\ \hline
  Ideal           & 26.9ms & 1.7ms \\ \hline
  ConsCPD         & 47.6ms & 1.7ms
  \end{tabular}
  \caption{Evaluation results for doubleAppendo}
  \label{tbl:doubleApp}
\end{table}

\includegraphics{graphs/doubleApp.pdf}

There is no significant difference in the execution time in the backward direction between different specialized versions (see table~\ref{tbl:doubleApp}).
However the execution time in the backward direction differs: the program generated by \ecce shows the same speedup as the ideal program, while conservative partial deduction slows the program down.
The reason behind the slowdown is the premature introduction of fresh variables which is significant for the input data of this size.

@@

\begin{table}
  \centering
  \begin{tabular}{c||c}
                   & [1..100] \\ \hline\hline
  Original         & 2.286ms  \\ \hline
  Ideal           & 4.623ms  \\ \hline
  Ideal removed   & 3.133ms  \\ \hline
  \ecce             & 1.701ms  \\ \hline
  ConsCPD          & 2.578ms
  \end{tabular}

@@

Surprising enough, the execution time of the ideal program is worse than of the original.
\ecce succeeds at improving the program performance, while conservative partial deduction worsens it a little.
\todo{why}

@@

The conservative partial deduction first unfolds those calls which are selected with the heuristic.
Since exploring boolean operations makes more sense, they are unfolded before recursive calls of \lstinline{eval$^o$}.
The way conservative partial deduction treats this program is the same as it treats the other implementation in which boolean operations are moved to the left, as shown in listing~\ref{eval:fst}.
This program is easier for CPD to transform which demonstrates how unequal is the behaviour of CPD for similar programs.

@@

The other way to implement boolean operations is via one basic boolean relation such as \lstinline{nand$^o$} which is, in turn, has a table-based implementation (see listing~\ref{not:nando}).
It will take several sequential unfoldings to derive that variables \lstinline{v} and \lstinline{w} should be \lstinline{^true} when considering a call \lstinline{and$^o$ v w ^true} implemented via a basic relation.
Conservative partial deduction drives the selected call until it derives useful substitutions for the variables involved while CPD with deterministic unfolding may fail to do so.

@@

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
                   & last    & plain  \\
  \hline\hline
  Original         & 0.695s  & 1.426s \\
  \hline
  \ecce             & 0.205s  & 0.264s \\
  \hline
  ConsCPD          & 0.132s  & 0.137s \\
  \end{tabular}

@@

The \lstinline{unify} relation was executed to find a unifier of two terms $f(X, X, g(Z, t))$ and $f(g(p, L), Y, Y)$.
The original \mk program fails to terminate on this goal in 30 seconds.
On this example, the most performant is the program generated by CPD (0.352 seconds) while the program generated by adding branching heuristic also fails to terminate in 30 seconds.
The conservative partial deduction shows some improvement with the residual program terminated within 1.947 seconds.
While driving this program, the conservative partial deduction does too much unfolding which negatively impacts the running time as compared to CPD.

The last test executed \lstinline{isPath} relation to search for 3 paths of length 7 in the graph with 20 vertices and 30 edges.
On this program, conservative partial deduction showed better transformation results than CPD, although the difference is not that drastic.

All evaluation results are presented in the table~\ref{tbl:unify}.
Each column corresponds to the relation being run as described above.
The row marked ``Original'' contains the running time of the original \mk relation before specialization, ``CPD'' and ``ConsCPD'' correspond to conjunctive and conservative partial deduction respectively while ``Branching'' is for the CPD modified with the branching heuristic.

\begin{table}
  \centering
  \begin{tabular}{c||c|c|c}
    \multirow{ 2}{*}{Terms} &
    f(X, a) & f(a \% b \% nil, c \% d \% nil, L) & f(X, X, g(Z, t))  \\
    \cline{2-4} &
    f(a, X) & f(X \% XS, YS, X \% ZS) & f(g(p, L), Y, Y)  \\
    \hline\hline
  Original         & $0.247*10^{-4}s$ & $>30s$  & $>30s$   \\
  \hline
  ConsCPD          & $0.084*10^{-4}s$ & $0.021s$ & $2.198s$  \\
  \hline
  \ecce             & $0.128*10^{-4}s$ & $0.026s$ & $1.340s$ \\
  \hline
  \end{tabular}

@@

% \begin{table}
%   \centering
%   \begin{tabular}{c||c|c|c||c}
%                    & \multicolumn{3}{c||}{unify} & isPath \\
%   \hline
%                    & first  & second & third     &        \\
%   \hline\hline
%   Original         & $0.247*10^{-4}s$ & $>30s$  & $>30s$  & $23.743s$ \\
%   \hline
%   ConsCPD          & $0.084*10^{-4}s$ & $0.021s$ & $2.198s$  & $1.786s$  \\
%   \hline
%   \ecce             & $0.128*10^{-4}s$ & $0.026s$ & $1.340s$  & $1.292s$ \\
%   \hline
%   \end{tabular}

@@

We compared two possible implementations of these typing rules.
The first one is obtained by unnesting of the functional program as described in~\cite{lozov2019relational}.
The second program is hand-written in \oc.
Each implementation has been transformed with conservative partial deducer and by \ecce.

We measured the time needed to generate 100 terms of type integer (averaged over 10 iterations).
\ecce significantly worsens the execution time for both implementations.
Conservative partial deduction improves the first implementation a little bit, while worsening the performance of the second implementation 3 times.

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
              & Implementation 1 & Implementation 2 \\ \hline\hline
  Original    & 0.464s           & 0.057s           \\ \hline
  ConsCPD     & 0.448s           & 0.154s           \\ \hline
  \ecce        & 1.518s           & 2.543s           \\
  \end{tabular}
